iSEE: Interactive SummarizedExperiment Explorer

### 2018-04-28 manuscript/main.tex
data set (Google says so) and missing 'as'

\begin{abstract}
Data exploration is critical to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
However, most existing tools for interactive visualization are limited to specific assays or analyses.
Here, we present the iSEE (Interactive SummarizedExperiment Explorer) software package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
iSEE is directly compatible with many existing R/Bioconductor packages for analyzing high-throughput biological data,
and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.

@@

Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the Shiny framework \citep{chang2017shiny} and exploits data structures from the open-source Bioconductor project \citep{gentleman2004bioconductor}, specifically the SummarizedExperiment class.
iSEE allows users to simultaneously visualize multiple aspects of a given dataset, including experimental data, metadata and analysis results.
Dynamic linking and point selection facilitate the flexible exploration of interactions between different data aspects.
Additional functionalities include code tracking, intelligent downsampling of large datasets, custom colour scale specification and tour construction.
We demonstrate the capabilities of iSEE by applying it to a diverse range of real datasets.

@@

\subsection*{Additional functionalities}
Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez. % do we need to cite them? at least ENSEMBL has updated citable papers
For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
Users can also include a bespoke step-by-step ``tour'' of their data set via the rintrojs package \cite{ganz2016rintrojs}, guiding the audience through an examination of the salient features in the data.

\section*{Use cases} % FM: used often in other software articles in F1000res
To demonstrate the functionalities and flexibility of the iSEE package, we applied our method to a number of publicly available datasets.
Details are provided in the tour subfolder of the GitHub repository for this manuscript (\url{https://github.com/LTLA/iSEE2018}), and for each dataset a combination of \_app.R and \_tour.txt is provided.
An instance of iSEE running on the respective datasets is provided as a stand-alone webpages, for demonstration purposes without the need for any software installation.

@@

\subsection*{Single-cell droplet-based RNA-seq}
% TODO: better title, or do we drop the allen section altogether (above)?
We also demonstrate the use of iSEE to explore a larger data set of 4,000 PBMC single-cell RNA-sequencing dataset from 10X Genomics \citep{zheng2017massively}.
% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?

@@

\subsection*{Bulk RNA-seq from TCGA}
We also demonstrate how iSEE seamlessly scales up with larger data sets as the TCGA RNA-seq collection of measurements for 23,368 genes across 7,706 tumor samples \citep{piccolo2015TCGA}.
% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?

@@

\section*{Conclusion}
iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
We provide a number of demonstrations at \url{https://github.com/LTLA/iSEE2018}, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells \citep{bodenmiller2012cytof}.
% TODO: repurpose the conclusion if we showcase the examples in earlier sections
% TODO: sell the idea that for every dataset it is possible to explore and showcase feats of interest. Something like...
We expect the functionalities and flexibility of the iSEE package to be essential for exploring and showcasing relevant features of any data set that can be represented in the form of a SummarizedExperiment object.
In particular, we expect public instances of iSEE to ideally accompany publications, as they enable authors to reproduce published figures and showcase salient aspects of their work using guided tours, while fostering hypothesis generation through transparent, interactive, reproducible, and efficient data analysis by the broader reader audience.

>>>>

\begin{abstract}
Data exploration is critical to the comprehension of large biological data sets generated by high-throughput assays such as sequencing.
However, most existing tools for interactive visualization are limited to specific assays or analyses.
Here, we present the iSEE (Interactive SummarizedExperiment Explorer) software package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
iSEE is directly compatible with many existing R/Bioconductor packages for analyzing high-throughput biological data,
and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics data sets.

@@

Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the Shiny framework \citep{chang2017shiny} and exploits data structures from the open-source Bioconductor project \citep{gentleman2004bioconductor}, specifically the SummarizedExperiment class.
iSEE allows users to simultaneously visualize multiple aspects of a given data set, including experimental data, metadata and analysis results.
Dynamic linking and point selection facilitate the flexible exploration of interactions between different data aspects.
Additional functionalities include code tracking, intelligent downsampling of large data sets, custom colour scale specification and tour construction.
We demonstrate the capabilities of iSEE by applying it to a diverse range of real data sets.

@@

\subsection*{Additional functionalities}
Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez. % do we need to cite them? at least ENSEMBL has updated citable papers
For large data sets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
Users can also include a bespoke step-by-step ``tour'' of their data set via the rintrojs package \cite{ganz2016rintrojs}, guiding the audience through an examination of the salient features in the data.

\section*{Use cases} % FM: used often in other software articles in F1000res
To demonstrate the functionalities and flexibility of the iSEE package, we applied our method to a number of publicly available data sets.
Details are provided in the tour subfolder of the GitHub repository for this manuscript (\url{https://github.com/LTLA/iSEE2018}), and for each data set a combination of \_app.R and \_tour.txt is provided.
An instance of iSEE running on the respective data sets is provided as a stand-alone webpages, for demonstration purposes without the need for any software installation.

@@

\subsection*{Single-cell droplet-based RNA-seq}
% TODO: better title, or do we drop the allen section altogether (above)?
We also demonstrate the use of iSEE to explore a larger data set of 4,000 PBMC single-cell RNA-sequencing data set from 10X Genomics \citep{zheng2017massively}.
% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?

@@

\subsection*{Bulk RNA-seq from TCGA}
We also demonstrate how iSEE seamlessly scales up with larger data sets such as the TCGA RNA-seq collection of measurements for 23,368 genes across 7,706 tumor samples \citep{piccolo2015TCGA}.
% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?

@@

\section*{Conclusion}
iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics data sets.
We provide a number of demonstrations at \url{https://github.com/LTLA/iSEE2018}, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing data set from 10X Genomics \citep{zheng2017massively} and a published mass cytometry data set consisting of over 170,000 cells \citep{bodenmiller2012cytof}.
% TODO: repurpose the conclusion if we showcase the examples in earlier sections
% TODO: sell the idea that for every data set it is possible to explore and showcase feats of interest. Something like...
We expect the functionalities and flexibility of the iSEE package to be essential for exploring and showcasing relevant features of any data set that can be represented in the form of a SummarizedExperiment object.
In particular, we expect public instances of iSEE to ideally accompany publications, as they enable authors to reproduce published figures and showcase salient aspects of their work using guided tours, while fostering hypothesis generation through transparent, interactive, reproducible, and efficient data analysis by the broader reader audience.
